An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-12-04 thesis/introduction-and-background.tex
updates
typos in thesis

Data from the host is first read into the
device's \emph{global memory}.
Similarly, data from the device is read back into the host
from global memory.
This is done in the host-code
using special functions provided by the CUDA API 
(e.g., \texttt{cudaMemcpy}).
Data transfer between the host and the device is
extremely slow relative to data access within the device.
This data transfer rate is limited by the bandwidth
of the PCI-e bus between the host and device.
All threads executing a kernel can
read and write to locations in global memory,
and it is the largest memory space that is
writable by the device.
For example, the current NVIDIA Tesla K20 accelerator
has about 4 Gigabytes of usable global memory.
Thread access to global memory has a long latency
and it is especially inefficient when
successive threads in a block access memory locations
that are far apart.
Data in global memory remains persistent
throughout the execution of the program.

@@

Despite these issues,
cyclic reduction remains an attractive algorithm
for GPUs,
for its low algorithmic complexity,
and lower work per step compared to PCR.
Much work has been done on
addressing these problems and
optimizing cyclic reduction performance on GPUs.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce the number and severity of bank conflicts,
and also to have better thread activity overall.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et al. \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focussed on exploiting the specific matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.

>>>>

Data from the host is first read into the
device's \emph{global memory}.
Similarly, data from the device is read back into the host
from global memory.
This is done in the host-code
using special functions provided by the CUDA API.
Data transfer between the host and the device is
extremely slow relative to data access within the device.
This data transfer rate is limited by the bandwidth
of the PCI-e bus between the host and device.
All threads executing a kernel can
read and write to locations in global memory,
and it is the largest memory space that is
writable by the device.
For example, the current NVIDIA Tesla K20 accelerator
has about 4 Gigabytes of usable global memory.
Thread access to global memory has a long latency
and it is especially inefficient when
successive threads in a block access memory locations
that are far apart.
Data in global memory remains persistent
throughout the execution of the program.

@@

Despite these issues,
cyclic reduction remains an attractive algorithm
for GPUs,
for its low algorithmic complexity,
and lower work per step compared to PCR.
Much work has been done on
addressing these problems and
optimizing cyclic reduction performance on GPUs.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce the number and severity of bank conflicts,
and also to have better thread activity overall.
G{\"o}ddeke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et al. \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focused on exploiting the specific matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.

### 2015-12-04 thesis/results.tex
thesis errata
typos in thesis

In this section we present a performance overview
of our NEATO solver
against a multi-threaded Intel MKL solver and
the CUSPARSE GPU solver
(\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
The MKL solver uses Gaussian elimination with partial pivoting,
and the CUSPARSE solver uses a combination of
Cyclic Reduction and Parallel Cyclic Reduction
as described by Zhang et al. \cite{Zhang2010FTS}.
These solvers represent the most straightforward way
to compute solutions for tridiagonal systems
and are highly optimized for performance on
underlying architectures.
Boundary conditions may prevent
the matrix from being symmetric and/or diagonally dominant,
precluding the use of more specialized tridiagonal solvers.

@@

\subsection{Strong and weak scaling}
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$.}
\label{fig:strong-scaling-256}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$.}
\label{fig:strong-scaling-512}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$ per process.}
\label{fig:weak-scaling-128}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$ per process.}
\label{fig:weak-scaling-256}
\end{center}
\end{figure}
%
Figures \ref{fig:strong-scaling-256} - \ref{fig:weak-scaling-256}
show the strong and weak scaling of the compact finite difference solver
for evaluating derivatives in all three coordinate directions.
For the strong scaling measurement,
we keep the problem size fixed
and increase the number of GPUs used to solve the problem.
For the weak scaling measurement,
we keep the problem size \emph{per GPU}
fixed, and increase the number of GPUs used.
We note good scaling for both 2-D and 3-D cases.
The strong scaling for 3-D problems is somewhat better
as the GPU is kept busy for all problem sizes.

>>>>

In this section we present a performance overview
of our NEATO solver
against a multi-threaded Intel MKL solver and
the CUSPARSE GPU solver
(\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
The MKL solver uses Gaussian elimination with partial pivoting,
and the CUSPARSE solver uses a combination of
Cyclic Reduction and Parallel Cyclic Reduction
as described by Zhang et al. \cite{Zhang2010FTS}.
These solvers represent the most straightforward way
to compute solutions for tridiagonal systems
and are highly optimized for performance on
underlying architectures.
boundary conditions may prevent
the matrix from being symmetric and/or diagonally dominant,
precluding the use of more specialized tridiagonal solvers.

@@

\subsection{Strong and weak scaling}
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$.}
\label{fig:strong-scaling-256}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$.}
\label{fig:strong-scaling-512}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$ per process.}
\label{fig:weak-scaling-128}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$ per process.}
\label{fig:weak-scaling-256}
\end{center}
\end{figure}
%
Figures \ref{fig:strong-scaling-256} - \ref{fig:weak-scaling-256}
show the strong and weak scaling of the compact finite difference solver
for evaluating derivatives in all three coordinate directions.
For the strong scaling measurement,
we keep the problem size fixed
and increase the number of GPUs used to solve the problem.
For the weak scaling measurement,
we keep the problem size \emph{per GPU}
fixed, and increase the number of GPUs used.
The strong scaling for larger problems is somewhat better
as the GPU is kept more busy.
